Sparse Function-space Representation \\ of Neural Networks

### 2023-05-17 paper/main.tex
small change

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that the neural network emits `dual' parameters which are the building blocks of a GP~\cite{csato2002sparse, adam2021dual, chang2020fast} . In contrast to previous work that utilizes subsets of training data \cite{immer2021scalable},
this parameterization allows capturing the contributions from {\em all} data points into a sparse representation, essential for predictive uncertainty. Crucially, the resulting GP directly predicts in the same space as the original trained neural network, with the benefit of avoiding the complexity introduced by working in weight-space and the notorious cubic complexity of vanilla GPs.
%\todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}
%, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs. 
Through the dual parameter formulation, we establish a connection between the neural network, full GPs, and a sparse approximation similar to sparse variational GPs~\cite{titsias2009variational}. We refer to our method as Sparse Function-space Representation (\our)---a sparse GP derived from a trained neural network. Moreover, this dual parameterization can be exploited to perform dual conditioning \citep{chang2022fantasizing}, \ie, an effective approach to condition on new incoming data into our model without need of retraning the model (see \cref{fig:teaser}).
%As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.
\looseness-2

@@

% Introduce the laplace GGN + shortcomings
\textbf{Function-space methods}
Function-space perspectives on uncertainty in neural networks often use a Laplace-GGN approximation \citep{daxberger2021laplace}, which linearizes a trained neural network around MAP weights and approximates the neural network's Hessian using the generalized Gauss--Newton approximation (GGN, \cite{botev2017practical}).
%\todo{Could modify the sentence a bit, maybe "The Laplace-GGN approximates the Hessian by using..."}
%The Hessian can be efficiently approximated using the generalized Gauss--Newton approximation . 
While efficient, it suffers from cubic scaling in parameter count, necessitating approximations like Kronecker factorisation \cite{martens2015optimizing, ritter2018kfac}. This linear model (linear in the weights) refines predictions and provides uncertainty estimates \citep{immer2021scalable}. It has a convenient interpretation as a GP \cite{immer2021scalable, khan2019approximate, maddox2021fast}.
%
However, the GP's cubic $\mathcal{O}(N^3)$ scaling with data points $N$ requires costly approximations, often resorting to using data subsets \cite{immer2021scalable}. In the GP community, sparse approximations have mitigated this scaling issue (\eg, \cite{titsias2009variational,hensman2013gaussian}), but combining neural network linearization with sparse methods remains unclear without resorting to subset methods or separately retraining a GP model \cite{ortega2023variational}. Our work addresses this by utilizing the GP's dual parameters \cite{csato2002sparse}, previously applied to non-conjugate likelihood models \cite{adam2021dual}.

% \todo{Could add something on subset methods here, for example 'which do not retain information from the full dataset' or something like that}.
%wang2019exact
%
%However, the GP introduces a cubic $\mathcal{O}(N^3)$ scaling in the number of data points $N$. Previous approaches apply crude approximations to alleviate the computational costs, \eg, by considering only a subset of the training data \cite{immer2021scalable}. In the GP community, the scaling problem is not seen as an issue any more due to efficient and stochastic methods for sparse approximations (\eg, \cite{hensman2013gaussian,wang2019exact}). However, it is not entirely clear how to combine the linearization of the neural network with sparse methods \cite{titsias2009variational} without falling back to crude subset methods. We tackle this problem by framing the problem in the dual parameters \cite{csato2002sparse} of the GP that has previously been used for non-conjugate likelihood models in GPs \cite{adam2021dual}.

@@

\textbf{Uncertainty quantification in RL}
A key challenge in RL is balancing the trade-off between exploration and exploitation \cite{sutton2018reinforcement}.
%That is, should an agent select actions that it knows will lead to high reward %(exploitation), or should it
%select new ones in hope to discover actions leading to higher reward (exploration).
One promising direction to balancing this trade-off is to model the uncertainty associated with a learned transition dynamics model and use it to guide exploration.
Prior work has taken an expectation over the dynamic model's posterior \cite{deisenroth2011pilco,kamtheDataEfficient2018,chuaDeepReinforcementLearning2018},
sampled from it (akin to Thompson sampling but referred to as posterior sampling in RL)
\todo{cite thopmson sampling paper dearden et al 1999}
\cite{osbandMoreEfficientReinforcement2013},
and used it to implement optimism in the face of uncertainty using upper confidence bounds \cite{curiEfficient2020}.
\todo{cite UCRL paper}
No single strategy has emerged as a go-to method and we highlight that these strategies are only as good as their uncertainty estimates.
Previous approaches have used GPs \cite{deisenroth2011pilco,kamtheDataEfficient2018},
ensembles of neural networks \cite{curiEfficient2020,chuaDeepReinforcementLearning2018}
and variational inference \cite{galImproving2016,houthooftVIME2017}. \todo{Could specify here how the previous methods have used these models}
However, each method has its pros and cons.
In this paper we present a method which combines some of the pros from NNs with the benefits of GPs.

%However, there has been little leveraging

@@

Using the Hessian of the approximate model, we arrive at the Laplace-GGN approximate posterior for $q(\vw)$ \todo{This is the first time q(w) is used, should define it}. 
Therefore, the Laplace-GGN linear model can be used to convert our weight space model to the function space by \todo{To be super clear even when reading this part out of context: could specify that using a GP with this mean and covariance},
\begin{equation} 
\label{eq-laplace-approx-function-space} 
% g(\vx) \sim \GP \left( \mu(\vx), \kappa(\vx, \vx') \right) \quad \text{with} \quad
  \mu(\vx) =  0 \quad \text{and} \quad
  \kappa(\vx, \vx')
  = \frac{1}{\delta} \Jac{\weights^*}{\vx} \, \JacT{\weights^*}{\vx'}, 
\end{equation}
where the kernel is the so-called Neural Tangent Kernel (NTK, \cite{jacot2018neural}). In \cref{eq-laplace-approx-function-space}, we restrict the function output to a single dimension for notational simplicity, yet this formulation can be readily extended to the multi-output scenario. With this kernel function, we can combine it with the data set $\dataset$ to construct the posterior. \todo{what is "it" here?}
%
\citet{khan2019approximate} took a similar approach, but instead of fitting the GP posterior to the actual $\vy$, they rely on a transformation of $\vy$. Similarly, \citet{immer2021improving} obtain the same covariance function but use a different mean because they rely on a first-order approximation of the neural network to form a Bayesian GLM model. Both approaches attempted to adjust the GP posterior mean function to ensure the predictions are from the neural network $f_{\vw^*}(\vx)$. 
Contrarily, our methodology intends to make predictions directly from the derived GP, thus evading such adjustments. Next, we will present the general formulation of the sparse functional representation of our trained neural network based on the non-sparse approximation.

>>>>

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that the neural network emits \todo{is 'emits' a clear enough expression here?}`dual' parameters which are the building blocks of a GP~\cite{csato2002sparse, adam2021dual, chang2020fast} . In contrast to previous work that utilizes subsets of training data \cite{immer2021scalable},
this parameterization allows capturing the contributions from {\em all} data points into a sparse representation, essential for predictive uncertainty. Crucially, the resulting GP directly predicts in the same space as the original trained neural network, with the benefit of avoiding the complexity introduced by working in weight-space and the notorious cubic complexity of vanilla GPs.
%\todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}
%, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs. 
Through the dual parameter formulation, we establish a connection between the neural network, full GPs, and a sparse approximation similar to sparse variational GPs~\cite{titsias2009variational}. We refer to our method as Sparse Function-space Representation (\our)---a sparse GP derived from a trained neural network. Moreover, this dual parameterization can be exploited to perform dual conditioning \citep{chang2022fantasizing}, \ie, an effective approach to condition on new incoming data into our model without need of retraning the model (see \cref{fig:teaser}).
%As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.
\looseness-2

@@

% Introduce the laplace GGN + shortcomings
\textbf{Function-space methods}
Function-space perspectives on uncertainty in neural networks often use a Laplace-GGN approximation \citep{daxberger2021laplace}, that linearize a trained neural network around MAP weights and approximates the neural network's Hessian using the generalized Gauss--Newton approximation (GGN, \cite{botev2017practical}). 
%\todo{Could modify the sentence a bit, maybe "The Laplace-GGN approximates the Hessian by using..."}
%The Hessian can be efficiently approximated using the generalized Gauss--Newton approximation . 
While efficient, it suffers from cubic scaling in parameter count, necessitating approximations like Kronecker factorisation \cite{martens2015optimizing, ritter2018kfac}. \todo{Reads a bit repetitive to start two sentences in a row by This, could maybe be combined in one.} This linear model (linear in the weights) refines predictions and provides uncertainty estimates \citep{immer2021scalable}. This model has a convenient interpretation as a GP \cite{immer2021scalable, khan2019approximate, maddox2021fast}.
%
However, the GP's cubic $\mathcal{O}(n^3)$ scaling with data points $n$ requires costly approximations, often resorting to using data subsets \cite{immer2021scalable}. In the GP community, sparse approximations have mitigated this scaling issue (\eg, \cite{titsias2009variational,hensman2013gaussian}), but combining neural network linearization with sparse methods remains \todo{Maybe unexplored instead of unclear?} unclear without resorting to subset methods or separately retraining a GP model \cite{ortega2023variational} \todo{Could add something on subset methods here, for example 'which do not retain information from the full dataset' or something like that}. Our work addresses this by utilizing the GP's dual parameters \cite{csato2002sparse}, previously applied to non-conjugate likelihood models \cite{adam2021dual}.

%wang2019exact
%
%However, the GP introduces a cubic $\mathcal{O}(N^3)$ scaling in the number of data points $N$. Previous approaches apply crude approximations to alleviate the computational costs, \eg, by considering only a subset of the training data \cite{immer2021scalable}. In the GP community, the scaling problem is not seen as an issue any more due to efficient and stochastic methods for sparse approximations (\eg, \cite{hensman2013gaussian,wang2019exact}). However, it is not entirely clear how to combine the linearization of the neural network with sparse methods \cite{titsias2009variational} without falling back to crude subset methods. We tackle this problem by framing the problem in the dual parameters \cite{csato2002sparse} of the GP that has previously been used for non-conjugate likelihood models in GPs \cite{adam2021dual}.

@@

\textbf{Uncertainty quantification in RL}
A key challenge in RL is balancing the trade-off between exploration and exploitation \cite{sutton2018reinforcement}.
%That is, should an agent select actions that it knows will lead to high reward %(exploitation), or should it
%select new ones in hope to discover actions leading to higher reward (exploration).
One promising direction to balancing this trade-off is to model the uncertainty associated with a learned transition dynamics model and use it to guide exploration.
Prior work has taken an expectation over the dynamic model's posterior \cite{deisenroth2011pilco,kamtheDataEfficient2018,chuaDeepReinforcementLearning2018},
sampled from it (akin to Thompson sampling but referred to as posterior sampling in RL)
\todo{cite thopmson sampling paper dearden et al 1999}
\cite{osbandMoreEfficientReinforcement2013},
and used it to implement optimism in the face of uncertainty using upper confidence bounds \cite{curiEfficient2020}.
\todo{cite UCRL paper}
No single strategy has emerged as a go-to method and we highlight that these strategies are only as good as their uncertainty estimates.
Previous approaches have used GPs \cite{deisenroth2011pilco,kamtheDataEfficient2018},
ensembles of neural networks \cite{curiEfficient2020,chuaDeepReinforcementLearning2018}
and variational inference \cite{galImproving2016,houthooftVIME2017}. \todo{Could specify here how the previous methods have used these models}
%However, there has been little leveraging

@@

Using the Hessian of the approximate model, we arrive at the Laplace-GGN approximate posterior for the weight space, $q(\vw)$. \todo{This is the first time q(w) is used, should define it}. 
Therefore, the Laplace-GGN linear model can be used to convert our weight space model to the function space by \todo{To be super clear even when reading this part out of context: could specify that using a GP with this mean and covariance},
\begin{equation} 
\label{eq-laplace-approx-function-space} 
% g(\vx) \sim \GP \left( \mu(\vx), \kappa(\vx, \vx') \right) \quad \text{with} \quad
  \mu(\vx) =  0 \quad \text{and} \quad
  \kappa(\vx, \vx')
  = \frac{1}{\delta} \Jac{\weights^*}{\vx} \, \JacT{\weights^*}{\vx'}, 
\end{equation}
where the kernel is the so-called Neural Tangent Kernel (NTK, \cite{jacot2018neural}). In \cref{eq-laplace-approx-function-space}, we restrict the function output to a single dimension for notational simplicity, yet this formulation can be readily extended to the multi-output scenario. With this kernel function, we can combine it with the data set $\dataset$ to construct the posterior. \todo{what is "it" here?}
%
\citet{khan2019approximate} took a similar approach, but instead of fitting the GP posterior to the actual $\vy$, they rely on a transformation of $\vy$. Similarly, \citet{immer2021improving} obtain the same covariance function but use a different mean because they rely on a first-order approximation of the neural network to form a Bayesian GLM model. Both approaches attempted to adjust the GP posterior mean function to ensure the predictions are from the neural network $f_{\vw^*}(\vx)$. 
Contrarily, our methodology intends to make predictions directly from the derived GP, thus evading such adjustments. Next, we will present the general formulation of the sparse functional representation of our trained neural network based on the non-sparse approximation.
